---
title: "Assignment_1"
author: "João Silva, Marco Pereira, Mariana Ribeiro"
date: "2026-03-04"
output: html_document
---
# Dataset description

The original data was acquired from **critically-ill patients** during hospital care at the Beth Israel Deaconess Medical Centre (Boston, MA, USA). 
**Two annotators** manually annotated individual breaths in each recording using the impedance respiratory signal. The **53 recordings** within the dataset, each of **8-minute** duration, each contain:

- Physiological signals include the photoplethysmogram **(PPG)**, impedance respiratory signal, and electrocardiogram (**ECG**). These are sampled at **125 Hz**.
- Physiological parameters, such as the heart rate (**HR**), respiratory rate (**RESP**), and blood oxygen saturation level (**SpO2**). These are sampled at 1 Hz.
- Fixed parameters, such as **age** and **gender**
- **Manual annotations of breaths**

# Installation and data loading

```{r}
# Install necessary packages
#install.packages("stream")
# Load necessary libraries
library(stream)
```

```{r}
# Download the dataset
url <- "https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_01_Numerics.csv?download"
download.file(url, destfile = "bidmc_01_Numerics.csv")
# Read the dataset
data <- read.csv("bidmc_01_Numerics.csv")
# Display the first few rows of the dataset
head(data)
```

```{r}
# Show the structure of the dataset
str(data)
```
```{r}
# Show summary statistics of the dataset
summary(data)
```
# Assignment 

Keep counters of equal-width grid-cells (base counters for micro-cluster definitions) of a 2-dimensional continuous data stream using different window models (landmark, sliding, weighted, fading).

```{r}
# Create two data stream from the dataset heart rate (HR) and respiratory rate (RESP)
stream1 <- DSD_Memory(data$HR)
stream2 <- DSD_Memory(data$RESP)
```

