---
title: "Assignment_1"
author: "João Silva, Marco Pereira, Mariana Ribeiro"
date: "2026-03-04"
output: html_document
---

# Dataset description

The original data was acquired from **critically-ill patients** during hospital care at the Beth Israel Deaconess Medical Centre (Boston, MA, USA). **Two annotators** manually annotated individual breaths in each recording using the impedance respiratory signal. The **53 recordings** within the dataset, each of **8-minute** duration, each contain:

-   Physiological signals include the photoplethysmogram **(PPG)**, impedance respiratory signal, and electrocardiogram (**ECG**). These are sampled at **125 Hz**.
-   Physiological parameters, such as the heart rate (**HR**), respiratory rate (**RESP**), and blood oxygen saturation level (**SpO2**). These are sampled at 1 Hz.
-   Fixed parameters, such as **age** and **gender**
-   **Manual annotations of breaths**

# Installation and data loading

```{r}
# Install necessary packages
#install.packages("stream")
#install.packages("animation")
# Load necessary libraries
library(stream)
library(animation)
```

```{r}
# Download the dataset
#url <- "https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_01_Numerics.csv?download"
#download.file(url, destfile = "bidmc_01_Numerics.csv")
# Read the dataset
data <- read.csv("bidmc_01_Numerics.csv")
# Display the first few rows of the dataset
head(data)
```

```{r}
# Show the structure of the dataset
str(data)
```

```{r}
# Show summary statistics of the dataset
summary(data)
```

# Assignment

Keep counters of equal-width grid-cells (base counters for micro-cluster definitions) of a 2-dimensional continuous data stream using different window models (landmark, sliding, weighted, fading).

```{r}
# Create a single 2-dimensional stream with heart rate (HR) and respiratory rate (RESP) directly from the file
stream_2d <- DSD_ReadCSV("bidmc_01_Numerics.csv", 
                           header = TRUE, 
                           sep = ",", 
                           take = c(2, 4)) # The 'take' parameter receives the numerical indices of the columns to be extracted.
```

### Parameter definition and grid setup

```{r}

# Time window parameters
w <- 100           # Objective window size (e.g., last 100 points)
eps <- 0.05        # Admissible ballast weight (maximum error threshold)
alpha <- eps^(1/w) # Weight factor calculation

# Define Equal-Width Grid-Cells 
# X-Axis: Heart Rate (HR) - typically between 0 and 200 bpm, bin size 10
breaks_hr <- seq(0, 200, by = 10)
# Y-Axis: Respiratory Rate (RESP) - typically between 0 and 60 breaths/min, bin size 2
breaks_resp <- seq(0, 60, by = 2)

# Com os parametros de cima a visualizacao aparece muito pequena, então para visualizar melhor é só necessário usar estes parametros:
#breaks_hr <- seq(78, 110, by = 5)
#breaks_resp <- seq(10, 36, by = 1)


n_bins_hr <- length(breaks_hr) - 1
n_bins_resp <- length(breaks_resp) - 1

```

### Initialization of the 4 window models

```{r}
# Each matrix represents the 2D grid where rows are HR and columns are RESP
grid_landmark <- matrix(0, nrow = n_bins_hr, ncol = n_bins_resp)
grid_sliding  <- matrix(0, nrow = n_bins_hr, ncol = n_bins_resp)
grid_weighted <- matrix(0, nrow = n_bins_hr, ncol = n_bins_resp)
grid_fading   <- matrix(0, nrow = n_bins_hr, ncol = n_bins_resp)

# Buffer list strictly for models with memory (Sliding and Weighted)
# It will store the (row, col) coordinates of the data points inside the window
buffer_history <- list()
```

### Stream processing (iterative step)

```{r}
itera <- 2000 # Simulated number of incoming data points

# Graphic window to exibit 4 plots in 2x2 format
par(mfrow = c(2, 2))

for (i in 1:itera) {
  
  # Read a single data point from the stream (Single Scan)
  pt <- get_points(stream_2d, n = 1)
  if (nrow(pt) == 0) break # Stop if stream is exhausted
  
  hr_val <- as.numeric(pt[1, 1])
  resp_val <- as.numeric(pt[1, 2])
  
  # Skip computation if sensor data is missing (NA)
  if (is.na(hr_val) || is.na(resp_val)) next
  
  # Find the correct grid-cell indices for the current data point
  idx_hr <- findInterval(hr_val, breaks_hr, all.inside = TRUE)
  idx_resp <- findInterval(resp_val, breaks_resp, all.inside = TRUE)
  
  current_coord <- c(idx_hr, idx_resp)
  
  # --------------------------------------------------------------------------
  # UPDATE: LANDMARK WINDOW (Accumulates forever)
  # --------------------------------------------------------------------------
  grid_landmark[idx_hr, idx_resp] <- grid_landmark[idx_hr, idx_resp] + 1
  
  # --------------------------------------------------------------------------
  # UPDATE: SLIDING WINDOW (Strictly the last 'w' elements)
  # --------------------------------------------------------------------------
  grid_sliding[idx_hr, idx_resp] <- grid_sliding[idx_hr, idx_resp] + 1
  
  # --------------------------------------------------------------------------
  # UPDATE: WEIGHTED SLIDING WINDOW (Decay + Explicit Removal)
  # --------------------------------------------------------------------------
  # Older data is less but still important
  grid_weighted <- grid_weighted * alpha
  grid_weighted[idx_hr, idx_resp] <- grid_weighted[idx_hr, idx_resp] + 1
  
  # --------------------------------------------------------------------------
  # UPDATE: FADING WINDOW (Memoryless Decay)
  # --------------------------------------------------------------------------
  # Uses all previous data to approximate a weighted window 
  # Formula: F_alpha,l(i) = c_li + alpha * F_alpha,l(i-1)
  grid_fading <- grid_fading * alpha
  grid_fading[idx_hr, idx_resp] <- grid_fading[idx_hr, idx_resp] + 1
  
  # --------------------------------------------------------------------------
  # MEMORY MANAGEMENT FOR SLIDING & WEIGHTED MODELS
  # --------------------------------------------------------------------------
  # Add the new coordinates to the end of the buffer
  buffer_history[[length(buffer_history) + 1]] <- current_coord
  
  # If the window exceeds its maximum size 'w', the oldest point must be forgotten
  if (length(buffer_history) > w) {
    
    # Retrieve the coordinates of the oldest point
    old_coord <- buffer_history[[1]]
    o_hr <- old_coord[1]
    o_resp <- old_coord[2]
    
    # Remove from Sliding Window (its weight was exactly 1)
    grid_sliding[o_hr, o_resp] <- grid_sliding[o_hr, o_resp] - 1
    
    # Remove from Weighted Sliding Window
    # The oldest point has decayed 'w' times, so its current residual weight is exactly alpha^w
    grid_weighted[o_hr, o_resp] <- grid_weighted[o_hr, o_resp] - (alpha^w)
    
    # Discard the oldest point from the buffer to maintain fixed memory size
    buffer_history[[1]] <- NULL
  }
  
  # COMEÇAR DESENVOLVIMENTO DA ANIMAÇÃO AQUI
  
}
```

```{r}
# ==============================================================================
# 2. DEFINIÇÃO DOS MODELOS DE JANELA E MICRO-CLUSTERS
# ==============================================================================
# A. Modelo Landmark (Marco)
dsc_landmark <- DSC_DStream(gridsize = 5, lambda = 0)

# B. Modelo Fading / Weighted (Desvanecimento)
dsc_fading <- DSC_DStream(gridsize = 5, lambda = 0.01)

# C. Modelo Sliding Window (Janela Deslizante estrita)
dsc_sliding <- DSC_Window(horizon = 100)

# D. Modelo Sliding Window Amortecida (Damped Window)
dsc_weighted_window <- DSC_Window(horizon = 100, lambda = 0.01)
```

# Animation of the evolution of micro-clusters (D-Stream Fading Model)

```{r}
# The reset_stream function returns to the beginning of the file.
reset_stream(stream_2d)

ani.options(interval = 0.2)

saveHTML({
  animate_cluster(
    dsc_fading,             
    stream_data, 
    horizon = 100,          
    n = 2000,               
    plot.args = list(
      type = "both",        
      grid = TRUE,          
      xlim = c(40, 140),  # Adjust the expected scale for HR
      ylim = c(0, 40),    # Adjust the expected scale for RESP
      xlab = "Heart Rate (HR)",
      ylab = "Respiratory Rate (RESP)",
      main = "Evolution 2D of Grid-Cells (D-Stream Fading Model)"
    )
  )
}, img.name = "stream_grid_animation", htmlfile = "evolucao_grelha.html")
```

# Explore and update other models

```{r}
# Sempre que necessitar de reavaliar o mesmo fluxo do início, faça reset:
reset_stream(stream_data)

update(dsc_sliding, stream_data, n = 1000)

cat("\n--- Centros da Janela Deslizante (Primeiros 6) ---\n")
print(head(get_centers(dsc_sliding)))

# Limpeza e fecho da ligação do ficheiro
close_stream(stream_data)
```

