---
title: "Assignment_2"
author: "João Silva, Marco Pereira, Mariana Ribeiro"
date: "2026-03-25"
output: pdf_document
---

# Dataset description

The original data was acquired from **critically-ill patients** during hospital care at the Beth Israel Deaconess Medical Centre (Boston, MA, USA). **Two annotators** manually annotated individual breaths in each recording using the impedance respiratory signal. The **53 recordings** within the dataset, each of **8-minute** duration, each contain:

-   Physiological signals include the photoplethysmogram **(PPG)**, impedance respiratory signal, and electrocardiogram (**ECG**). These are sampled at **125 Hz**.
-   Physiological parameters, such as the heart rate (**HR**), respiratory rate (**RESP**), and blood oxygen saturation level (**SpO2**). These are sampled at 1 Hz.
-   Fixed parameters, such as **age** and **gender**
-   **Manual annotations of breaths**

# Load packages

```{r}
# Install packages
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("mclust")) install.packages("mclust")

# Load libraries
library(tidyverse)
library(mclust)
```

# Exercise

Using all the files available in https://physionet.org/content/bidmc/1.0.0/:

1. Create an **object** containing the **mean HR and age** for **each subject**.
2. **Plot the histogram** of all the **HR means** with the respective **density curve**.
3. Apply the **expectation-maximization algorithm with 2 and 3 latent classes**.
4. **Plot the densities** for **each** of the **latent classes**. 
5. **Plot** the **mean** HR as a **function** of each subject's **age** in which **each point** has a **color** corresponding to the **class to which it belongs**.
6. **Generate 1000 data points** at random according to a **single distribution fitted to the original data**.
7. **Generate 1000 data points** at random according to a **mixture of distributions** fitted using **one of the previous EM** computed in 3.

The submitted code file should include **comments** to improve readability.

Notes:
- the filenames vary between bidmc_01 to bidmc_53
- the **bidmc_XX_Numerics.csv** contains four columns with the physiological parameters, such as the heart rate (**HR**), respiratory rate (RESP), and blood oxygen saturation level (SpO2). These are sampled at 1 Hz.
- the **bidmc_XX_Fix.txt** contains the values of Fixed variables, such as **age** and gender.
 - the link to download the files are in the following format - "https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_01_Numerics.csv?download" and "https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_01_Fix.txt"
- do not also forget to use the **functions density and EM** already **available in R**.

# 1. Download data and create an data frame containing the mean HR and age for each subject.

```{r}
# Create empty data frame to store mean HR and age for each subject
data_summary <- data.frame(subject = character(),
                           mean_HR = numeric(),
                           age = numeric(),
                           stringsAsFactors = FALSE)
# Loop through each subject's files
for (i in 1:53) {
  # Create file names
  numerics_file <- paste0("https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_", sprintf("%02d", i), "_Numerics.csv?download")
  fix_file <- paste0("https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_", sprintf("%02d", i), "_Fix.txt?download")
  # Read the numerics data
  numerics_data <- read.csv(numerics_file)
  # Read the fixed data
  fix_data <- readLines(fix_file, warn = FALSE)
  # Calculate mean HR
  mean_HR <- mean(numerics_data$HR, na.rm = TRUE)
  # Get age from fixed data
  age <- as.integer(sub(".*?([0-9]+).*", "\\1", fix_data[6]))  # Assuming age is in the sixth row
  # Append to the summary data frame
  data_summary <- rbind(data_summary, data.frame(subject = paste0("bidmc_",
                                                          sprintf("%02d", i)),
                                                 mean_HR = mean_HR,
                                                 age = age))
}
```
Subject 18 age is missing in the original file.

# 2. Plot the histogram of all the HR means with the respective density curve.

```{r}
# Plot histogram of mean HR with density curve
ggplot(data_summary, aes(x = mean_HR)) +
  geom_histogram(aes(y = ..density..), bins = 10, fill = "lightblue", color = "black") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Histogram of Mean HR with Density Curve",
       x = "Mean Heart Rate (HR)",
       y = "Density") +
  theme_minimal()
```


# 3. Apply the expectation-maximization algorithm with 2 and 3 latent classes.

```{r}
# Apply EM algorithm with 2 latent classes
em_2 <- Mclust(data_summary$mean_HR, G = 2)
# Apply EM algorithm with 3 latent classes
em_3 <- Mclust(data_summary$mean_HR, G = 3)
```


# 4. Plot the densities for each of the latent classes. 

# 5. Plot the mean HR as a function of each subject's age in which each point has a color corresponding to the class to which it belongs.

# 6. Generate 1000 data points at random according to a single distribution fitted to the original data.

# 7. Generate 1000 data points at random according to a mixture of distributions fitted using one of the previous EM computed in 3.