---
title: "Assignment_1"
author: "João Silva, Marco Pereira, Mariana Ribeiro"
date: "2026-03-04"
output: html_document
---

# Dataset description

The original data was acquired from **critically-ill patients** during hospital care at the Beth Israel Deaconess Medical Centre (Boston, MA, USA). **Two annotators** manually annotated individual breaths in each recording using the impedance respiratory signal. The **53 recordings** within the dataset, each of **8-minute** duration, each contain:

-   Physiological signals include the photoplethysmogram **(PPG)**, impedance respiratory signal, and electrocardiogram (**ECG**). These are sampled at **125 Hz**.
-   Physiological parameters, such as the heart rate (**HR**), respiratory rate (**RESP**), and blood oxygen saturation level (**SpO2**). These are sampled at 1 Hz.
-   Fixed parameters, such as **age** and **gender**
-   **Manual annotations of breaths**

# Load packages and dataset

```{r}
# Install packages
#install.packages("stream")
#install.packages("animation")

# Load libraries
library(stream)
library(animation)


library(ggplot2)
library(dplyr)
```

Dataset here is loaded **only for exploration and visualization purposes**. The stream will read the file sequentially without loading the entire dataset into memory.

```{r}
# Download the dataset
#url <- "https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_01_Numerics.csv?download"
#download.file(url, destfile = "bidmc_01_Numerics.csv")

# The stream reads the file sequentially without loading the entire dataset into memory
data <- read.csv("bidmc_01_Numerics.csv")

# Display the first few rows of the dataset
head(data)
```

```{r}
# Show the structure of the dataset
str(data)
```

```{r}
# Show summary statistics of the dataset
summary(data)
```

# Assignment

Keep counters of equal-width grid-cells (base counters for micro-cluster definitions) of a 2-dimensional continuous data stream using different window models (landmark, sliding, weighted, fading).

Two approaches are implemented:

1)  Manual implementation of grid-cell counters
    -   Landmark window
    -   Sliding window
    -   Weighted window
    -   Fading window
2)  Use of built-in models from the 'stream' package
    -   D-Stream algorithm
    -   Sliding window model

The data stream consists of two physiological variables: - Heart Rate (HR) and Respiratory Rate (RESP).

## 1)  Manual implementation of grid-cell counters

```{r}
# Create a single 2-dimensional stream with HR and RESP directly from the file
  # The stream reads the file sequentially without loading the entire dataset into memory
stream_2d <- DSD_ReadCSV("bidmc_01_Numerics.csv", 
                           header = TRUE, 
                           sep = ",", 
                           take = c(2, 4)) # HR and RESP column indices
```

### Parameter definition and grid setup

```{r}
# Time window parameters
w <- 100           # Sliding window size
eps <- 0.05         # Error tolerance
alpha <- eps^(1/w)  # Exponential decay factor

iterations <- 481  # Number of stream points processed

# Define Equal-Width Grid-Cells 
# X-Axis: HR - typically between 0 and 200 bpm, bin size 10
breaks_hr <- seq(88, 95, by = 1)
# Y-Axis: RESP - typically between 0 and 60 breaths/min, bin size 2
breaks_resp <- seq(20, 26, by = 1)

n_bins_hr <- length(breaks_hr) - 1 # Number of bins for HR
n_bins_resp <- length(breaks_resp) - 1 # Number of bins for RESP

```

### Initialization of grid counters (manual implementation)

```{r}
# Each matrix represents the 2D grid where rows are HR and columns are RESP
grid_landmark <- matrix(0, nrow = n_bins_hr, ncol = n_bins_resp)
grid_sliding  <- matrix(0, nrow = n_bins_hr, ncol = n_bins_resp)
grid_weighted <- matrix(0, nrow = n_bins_hr, ncol = n_bins_resp)
grid_fading   <- matrix(0, nrow = n_bins_hr, ncol = n_bins_resp)

# Buffer storing coordinates of the last w observations
# It will store the (row, col) coordinates of the data points inside the window
buffer_history <- list()
```

### Stream processing (iterative step) - MANUAL APPROACH

```{r}
# Animation of the evolution of grid counters for the four window models
saveGIF({

par(mfrow=c(2,2)) # Set up a 2x2 plotting area for the four grid visualizations

# Iterate through the stream points
for(i in 1:iterations){

  # Retrieve one point from the stream
  pt <- get_points(stream_2d,1)

  #Definition of features
  # column 1 -  HR, column 2 - RESP
  hr_val   <- pt[[1, 1]] 
  resp_val <- pt[[1, 2]]
  
  # If the stream has ended, break the loop
  if(nrow(pt)==0) break

  # Skip missing sensor readings
  if(is.na(hr_val) || is.na(resp_val)) next


  # --------------------------------------------------------
  # Map the observation to a grid cell
  # --------------------------------------------------------

  idx_hr <- findInterval(hr_val,breaks_hr,all.inside=TRUE)
  idx_resp <- findInterval(resp_val,breaks_resp,all.inside=TRUE)

  coord <- c(idx_hr,idx_resp)
  
  # --------------------------------------------------------
  # LANDMARK WINDOW
  # Counts all observations since the beginning
  # --------------------------------------------------------

  grid_landmark[idx_hr,idx_resp] <-
    grid_landmark[idx_hr,idx_resp] + 1


  # --------------------------------------------------------
  # SLIDING WINDOW
  # Keeps only the last w observations
  # --------------------------------------------------------

  grid_sliding[idx_hr,idx_resp] <-
    grid_sliding[idx_hr,idx_resp] + 1


  # --------------------------------------------------------
  # WEIGHTED WINDOW
  # Applies exponential decay to older observations
  # --------------------------------------------------------

  grid_weighted <- grid_weighted * alpha

  grid_weighted[idx_hr,idx_resp] <-
    grid_weighted[idx_hr,idx_resp] + 1

  
  # --------------------------------------------------------
  # FADING WINDOW
  # Memoryless exponential decay model
  # --------------------------------------------------------

  grid_fading <- grid_fading * alpha

  grid_fading[idx_hr,idx_resp] <-
    grid_fading[idx_hr,idx_resp] + 1


  # --------------------------------------------------------
  # BUFFER MANAGEMENT (for sliding window)
  # --------------------------------------------------------
  buffer_history[[length(buffer_history)+1]] <- coord

  if(length(buffer_history) > w){

    old <- buffer_history[[1]]

    grid_sliding[old[1],old[2]] <-
      grid_sliding[old[1],old[2]] - 1

    buffer_history[[1]] <- NULL
  }

  # --------------------------------------------------------
  # VISUALIZATION
  # --------------------------------------------------------

  image(x = breaks_hr, y = breaks_resp, z = grid_landmark,
        main="Landmark Window",
        xlab="Heart Rate (BPM)",
        ylab="Respiration Rate")

  image(x = breaks_hr, y = breaks_resp, z = grid_sliding,
        main="Sliding Window",
        xlab="Heart Rate (BPM)",
        ylab="Respiration Rate")

  image(x = breaks_hr, y = breaks_resp, z = grid_weighted,
        main="Weighted Window",
        xlab="Heart Rate (BPM)",
        ylab="Respiration Rate")

  image(x = breaks_hr, y = breaks_resp, z = grid_fading,
        main="Fading Window",
        xlab="Heart Rate (BPM)",
        ylab="Respiration Rate")

}

}, interval=0.2)

```

```{r}
#Static graph

# Reset stream (optional?)
reset_stream(stream_2d)

# Make a repository from captured points
captured_points <- data.frame()

for(i in 1:iterations) {
  pt <- get_points(stream_2d, 1)
  if(nrow(pt) == 0) break
  
  hr_val <- pt[[1, 1]]
  resp_val <- pt[[1, 2]]
  
  if(is.na(hr_val) || is.na(resp_val)) next
  
  # Fill in dataframe
  captured_points <- rbind(captured_points, data.frame(time = i, hr = hr_val, resp = resp_val))
}

# Frequency per cell (Grelha)
captured_points <- captured_points %>%
  mutate(
    bin_hr = breaks_hr[findInterval(hr, breaks_hr, all.inside = TRUE)],
    bin_resp = breaks_resp[findInterval(resp, breaks_resp, all.inside = TRUE)]
  ) %>%
  group_by(bin_hr, bin_resp) %>%
  mutate(count = n()) %>% # Number of times per cell
  ungroup()

# 1. Counting occurrences per cell (same bins)
plot_data <- captured_points %>%
  group_by(bin_hr, bin_resp) %>%
  mutate(frequency = n()) %>% 
  ungroup()

ggplot(plot_data, aes(x = hr, y = resp)) +
  # 1. Pathway
  geom_path(color = "gray70", linetype = "dashed", alpha = 0.5) +
  
  # 2. Points
  # 'size = frequency' bigger points with more ocurrences
  geom_point(aes(size = frequency, color = frequency), alpha = 1) +
  
  # 3. Scaling and color
  scale_size_continuous(range = c(2, 20)) +
  scale_color_viridis_c() +

  # 4. Legend
 guides(
    size = guide_legend(title = "Ocurrences"),
    color = guide_legend(title = "Ocurrences")
  ) +
  
  theme_minimal() +
  labs(
    title = "Heart and Respiratory Rate Stream: The Pathway",
    
    x = "Heart Rate (BPM)",
    y = "Respiration Rate (CPM)"
  ) +
  theme(legend.position = "right")
```
### Reseting stream before 2nd approach
```{r}
reset_stream(stream_2d)
```

## 2)  Use of built-in models from the 'stream' package

### Stream processing (iterative step) - USING BUILT-IN MODELS FROM 'stream' PACKAGE

```{r}
# Lambda value for fading models (D-Stream with decay and weighted sliding window)

lambda_value <- -log(eps) / w

# These models implement micro-cluster algorithms internally.

# Landmark-like behaviour (no fading)
dsc_landmark <- DSC_DStream(
  gridsize = 1,
  lambda = 0
)

# Fading model (D-Stream with decay)
dsc_fading <- DSC_DStream(
  gridsize = 1,
  lambda = lambda_value
)

# Strict sliding window model
dsc_sliding <- DSC_Window(
  horizon = 100
)

# Weighted sliding window model
dsc_weighted <- DSC_Window(
  horizon = 100,
  lambda = lambda_value
)
```


### Animation of the evolution of micro-clusters (D-Stream Fading Model)

```{r}
ani.options(interval = 0.2)

saveHTML({
  animate_cluster(
    dsc_fading,             
    stream_2d, 
    horizon = 100,          
    n = 2000,               
    plot.args = list(
      type = "both",        
      grid = TRUE,          
      xlim = c(40, 140),  # Adjust the expected scale for HR
      ylim = c(0, 40),    # Adjust the expected scale for RESP
      xlab = "Heart Rate (HR)",
      ylab = "Respiratory Rate (RESP)",
      main = "2D Grid Evolution (D-Stream Fading Model)"
    )
  )
},
img.name = "stream_grid_animation",
htmlfile = "evolucao_grelha.html")
```

```{r}
reset_stream(stream_2d)
```

### Animation of comparative models
```{r}
# ---------------------------------------------------------
# Comparative Memory Models
# ---------------------------------------------------------

# Parameters for Phase-Space and Windows
w <- 100           # Sliding window size
eps <- 0.05         # Error tolerance
alpha_fading <- eps^(1/w)  # Exponential decay factor
horizon_linear <- 100  # Cut-off horizon for Linear Weighted Window
history_hr <- numeric() # Buffer to store HR values for lag calculation

history_hr     <- numeric() # Buffer to store Heart Rate values
history_resp     <- numeric() # Buffer to store RR Interval values

saveGIF({
  # Setup a 3x2 grid layout
  # mar: margins, oma: outer margins for the main title
  par(mfrow = c(3, 2), mar = c(4, 4, 3, 1), oma = c(0, 0, 2, 0))
  
  for(i in 1:iterations) {
    # 1. Data Ingestion: Retrieve 1 point from the stream
    pt <- get_points(stream_2d, 1)
    if(nrow(pt) == 0) break
    
    hr_val  <- pt[[1, 1]]
    resp_val  <- pt[[1, 2]] 
    
    # Skip missing sensor readings to maintain data integrity
    if(is.na(hr_val) || is.na(resp_val)) next
    
    # 2. History Management: Update the buffers
    history_hr <- c(history_hr, hr_val)
    history_resp <- c(history_resp, resp_val)
    n_pts      <- length(history_hr)
    
    # Define fixed axis limits for visual consistency across frames
    lim_hr <- range(breaks_hr) 
    lim_resp <- range(breaks_resp) 

    # --- Visualizations ---

    # PLOT 1: Real-Time State
    # Represents the "Now": the most recent observation in the phase space
    plot(hr_val, resp_val, pch = 4, col = "red", cex = 2.5, lwd = 2,
         main = "1. Real Time (Current State)", xlab = "HR (bpm)", ylab = "RR (s)",
         xlim = lim_hr, ylim = lim_resp)
    grid()

    # PLOT 2: Dashboard Info
    # Textual representation of the current stream iteration
    plot.new()
    text(0.5, 0.7, paste("Observation:", i), cex = 1.8, font = 2)
    text(0.5, 0.4, paste("HR:", round(hr_val, 1), "bpm"), cex = 1.2)
    text(0.5, 0.2, paste("RR:", round(resp_val, 3), "s"), cex = 1.2)

    # PLOT 3: Landmark Window (Infinite Memory)
    # Cumulative plot of all historical points with transparency to show density
    plot(history_hr, history_resp, pch = 16, col = rgb(0.2, 0.2, 0.2, 0.1), 
         main = "2. Landmark (Infinite Memory)", xlab = "HR", ylab = "RR",
         xlim = lim_hr, ylim = lim_resp)

    # PLOT 4: Sliding Window (Finite Discrete Memory)
    # Only the last 'w' points are kept; older data is completely discarded
    idx_s <- max(1, n_pts - w + 1)
    plot(history_hr[idx_s:n_pts], history_resp[idx_s:n_pts], pch = 16,
         col = "blue", main = paste("3. Sliding (w =", w, ")"),
         xlab = "HR", ylab = "RR", xlim = lim_hr, ylim = lim_resp)

    # PLOT 5: Fading Window (Exponential Decay)
    # Points fade out smoothly; the weight of history decreases geometrically
    weights_fading <- alpha_fading^((n_pts:1) - 1)
    plot(history_hr, history_resp, pch = 16, col = rgb(1, 0.5, 0, alpha = weights_fading), 
         main = "4. Fading (Exp. Decay)", xlab = "HR", ylab = "RR",
         xlim = lim_hr, ylim = lim_resp)

    # PLOT 6: Weighted Window (Linear Decay)
    # Points lose weight linearly until they reach the 'horizon' and disappear
    weights_linear <- pmax(0, (horizon_linear - (n_pts:1 - 1)) / horizon_linear)
    plot(history_hr, history_resp, pch = 16, col = rgb(0.5, 0, 0.5, alpha = weights_linear), 
         main = "5. Weighted (Linear)", xlab = "HR", ylab = "RR",
         xlim = lim_hr, ylim = lim_resp)
  }
  
  # Global title for the animation across all subplots
  title("Stream Memory Models Comparison: HR vs RR Correlation", outer = TRUE)
  
}, movie.name = "hr_rr_memory_analysis.gif", interval = 0.1)
```

The visualization contrasts infinite memory (Landmark) against discrete (Sliding) and fading (Exponential/Linear) models, demonstrating the trade-off between historical density and temporal reactivity.
The Landmark model preserves the global trajectory at the cost of high memory consumption, whereas the Sliding window provides a computationally efficient but discrete snapshot of the last 100 iterations (in this case). To address the loss of temporal nuance in uniform weighting, the Fading and Weighted models were implemented; the former employs an exponential decay to maintain a smooth historical trace, while the latter uses a linear gradient to phase out the relevance of aging data points, exibiting similar behaviour.

# Close stream connection
```{r}
close_stream(stream_2d)
```
