---
title: "Assignment_2"
author: "João Silva, Marco Pereira, Mariana Ribeiro"
date: "2026-03-26"
output: pdf_document
---

# Dataset description

The original data was acquired from **critically-ill patients** during hospital care at the Beth Israel Deaconess Medical Centre (Boston, MA, USA). **Two annotators** manually annotated individual breaths in each recording using the impedance respiratory signal. The **53 recordings** within the dataset, each of **8-minute** duration, each contain:

-   Physiological signals include the photoplethysmogram **(PPG)**, impedance respiratory signal, and electrocardiogram (**ECG**). These are sampled at **125 Hz**.
-   Physiological parameters, such as the heart rate (**HR**), respiratory rate (**RESP**), and blood oxygen saturation level (**SpO2**). These are sampled at 1 Hz.
-   Fixed parameters, such as **age** and **gender**
-   **Manual annotations of breaths**

# Load packages

```{r, echo=FALSE}

# Install packages
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("mclust")) install.packages("mclust")

# Load libraries
library(tidyverse)
library(mclust)
```

# Exercise

Using all the files available in <https://physionet.org/content/bidmc/1.0.0/>:

1.  Create an **object** containing the **mean HR and age** for **each subject**.
2.  **Plot the histogram** of all the **HR means** with the respective **density curve**.
3.  Apply the **expectation-maximization algorithm with 2 and 3 latent classes**.
4.  **Plot the densities** for **each** of the **latent classes**.
5.  **Plot** the **mean** HR as a **function** of each subject's **age** in which **each point** has a **color** corresponding to the **class to which it belongs**.
6.  **Generate 1000 data points** at random according to a **single distribution fitted to the original data**.
7.  **Generate 1000 data points** at random according to a **mixture of distributions** fitted using **one of the previous EM** computed in 3.

The submitted code file should include **comments** to improve readability.

Notes: - the filenames vary between bidmc_01 to bidmc_53 - the **bidmc_XX_Numerics.csv** contains four columns with the physiological parameters, such as the heart rate (**HR**), respiratory rate (RESP), and blood oxygen saturation level (SpO2). These are sampled at 1 Hz. - the **bidmc_XX_Fix.txt** contains the values of Fixed variables, such as **age** and gender. - the link to download the files are in the following format - "<https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_01_Numerics.csv?download>" and "<https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_01_Fix.txt>" - do not also forget to use the **functions density and EM** already **available in R**.

# 1. Download data and create a data frame containing the mean HR and age for each subject.

```{r}
# Create empty data frame to store mean HR and age for each subject
data_summary <- data.frame(subject = character(),
                           mean_HR = numeric(),
                           age = numeric(),
                           stringsAsFactors = FALSE)
# Loop through each subject's files
for (i in 1:53) {
  # Create file links for numerics and fixed data
  numerics_file <- paste0("https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_", sprintf("%02d", i), "_Numerics.csv?download")
  fix_file <- paste0("https://physionet.org/files/bidmc/1.0.0/bidmc_csv/bidmc_", sprintf("%02d", i), "_Fix.txt?download")
  
  # Read the numerics data
  numerics_data <- read.csv(numerics_file)
  
  # Read the fixed data
  fix_data <- readLines(fix_file, warn = FALSE)
  
  # Calculate mean HR
  mean_HR <- mean(numerics_data$HR, na.rm = TRUE)
  
  # Get age from fixed data
  age <- as.integer(sub(".*?([0-9]+).*", "\\1", fix_data[6]))  # Assuming age is in the sixth row
  
  # Append to the summary data frame
  data_summary <- rbind(data_summary, data.frame(subject = paste0("bidmc_",
                                                        sprintf("%02d", i)),
                                                        mean_HR = mean_HR,
                                                        age = age))
}
```

**Note:** Subject 18 age is missing in the original file!

# 2. Plot the histogram of all the HR means with the respective density curve.

## Base R

```{r}
# Plot histogram of mean HR with density curve using base R
hist(data_summary$mean_HR, col = "lightblue",main = "Histogram of Mean HR with Density Curve", xlab = "Mean Heart Rate (HR)", ylab = "Density", freq = FALSE, ylim = c(0, 0.04)) # freq = FALSE to plot density instead of counts
lines(density(data_summary$mean_HR), col = "red", lwd = 2) # Add density curve
```

## Ggplot2

```{r}
# Plot histogram of mean HR with density curve
ggplot(data_summary, aes(x = mean_HR)) + 
  geom_histogram(aes(y = after_stat(density)), bins = 10, fill = "lightblue", color = "black") + # After stat(density) to plot density on the y-axis. The bars will represent the density instead of counts.
  geom_density(color = "red", linewidth = 1) + # Add density curve
  labs(title = "Histogram of Mean HR with Density Curve",
       x = "Mean Heart Rate (HR)",
       y = "Density") +
  theme_minimal()
```

# 3. Apply the expectation-maximization algorithm with 2 and 3 latent classes.

## Manual implementation soft

### 2 latent classes

```{r}
# iteration 0
m <- 2 # number of latent classes
eps <- 0.01 # convergence error
itera <- 0 # iteration counter
maxitera <- 100 # maximum number of iterations
set.seed(523) # for reproducibility
param.mean <- sample(data_summary$mean_HR, m) # initial mean values for the latent classes, randomly sampled from the data
param.sd <- rep(sd(data_summary$mean_HR) / m, m) # initial standard deviation values for the latent classes, set to a fraction of the overall standard deviation of the data

for (itera in 0:maxitera) { # EM algorithm loop
  names(param.mean) <- paste("mean",1:m) # assign names to the mean parameters for better readability
  names(param.sd) <- paste("sd",1:m) # assign names to the standard deviation parameters for better readability
  print(param.mean) # print the current mean parameters for each latent class
  print(param.sd) # print the current standard deviation parameters for each latent class

  ## Expectation-Maximization - E-Step

  xx <- data_summary$mean_HR # data points (mean HR for each subject)
  densities <- sapply(1:m, function(i){dnorm(xx,mean=param.mean[i],sd=param.sd[i])}) # calculate the density of each data point under each latent class distribution
  relevance <- t(apply(densities, 1, function(l){l/sum(l)})) # calculate the relevance (responsibility) of each latent class for each data point by normalizing the densities

  ## Expectation-Maximization - M-Step soft

  param.mean.new <- sapply(1:m, function(j){sum(data_summary$mean_HR*relevance[,j])/sum(relevance[,j])}) # update the mean parameters for each latent class by calculating the weighted average of the data points, where the weights are given by the relevance of each latent class for each data point
  param.sd.new <- sapply(1:m, function(j){sqrt(sum(relevance[,j]*(data_summary$mean_HR-param.mean[j])^2)/sum(relevance[,j]))}) # update the standard deviation parameters for each latent class by calculating the weighted standard deviation of the data points, where the weights are given by the relevance of each latent class for each data point

  ## Expectation-Maximization - Convergence

  cat("Iteration",itera,"evolution:",dif <- sum(abs(param.mean.new - param.mean)), "\n") # calculate the difference between the new and old mean parameters to check for convergence
  if (dif < eps) break # if the difference is less than the convergence threshold, exit the loop
  param.mean <- param.mean.new # update the mean parameters for the next iteration
  param.sd <- param.sd.new # update the standard deviation parameters for the next iteration
}
```

### 3 latent classes

```{r}
# iteration 0
m <- 3 # number of latent classes
eps <- 0.01 # convergence error
itera <- 0 # iteration counter
maxitera <- 100 # maximum number of iterations
set.seed(523) # for reproducibility
param.mean <- sample(data_summary$mean_HR, m) # initial mean values for the latent classes, randomly sampled from the data
param.sd <- rep(sd(data_summary$mean_HR) / m, m) # initial standard deviation values for the latent classes, set to a fraction of the overall standard deviation of the data

for (itera in 0:maxitera) { # EM algorithm loop
  names(param.mean) <- paste("mean",1:m) # assign names to the mean parameters for better readability
  names(param.sd) <- paste("sd",1:m) # assign names to the standard deviation parameters for better readability
  print(param.mean) # print the current mean parameters for each latent class
  print(param.sd) # print the current standard deviation parameters for each latent class

  ## Expectation-Maximization - E-Step

  xx <- data_summary$mean_HR # data points (mean HR for each subject)
  densities <- sapply(1:m, function(i){dnorm(xx,mean=param.mean[i],sd=param.sd[i])}) # calculate the density of each data point under each latent class distribution
  relevance <- t(apply(densities, 1, function(l){l/sum(l)})) # calculate the relevance (responsibility) of each latent class for each data point by normalizing the densities

  ## Expectation-Maximization - M-Step soft

  param.mean.new <- sapply(1:m, function(j){sum(data_summary$mean_HR*relevance[,j])/sum(relevance[,j])}) # update the mean parameters for each latent class by calculating the weighted average of the data points, where the weights are given by the relevance of each latent class for each data point
  param.sd.new <- sapply(1:m, function(j){sqrt(sum(relevance[,j]*(data_summary$mean_HR-param.mean[j])^2)/sum(relevance[,j]))}) # update the standard deviation parameters for each latent class by calculating the weighted standard deviation of the data points, where the weights are given by the relevance of each latent class for each data point

  ## Expectation-Maximization - Convergence

  cat("Iteration",itera,"evolution:",dif <- sum(abs(param.mean.new - param.mean)), "\n") # calculate the difference between the new and old mean parameters to check for convergence
  if (dif < eps) break # if the difference is less than the convergence threshold, exit the loop
  param.mean <- param.mean.new # update the mean parameters for the next iteration
  param.sd <- param.sd.new # update the standard deviation parameters for the next iteration
}
```

## Manual implementation crisp

### 2 latent classes

```{r}
# iteration 0
m <- 2 # number of latent classes
eps <- 0.01 # convergence error
itera <- 0 # iteration counter
maxitera <- 100 # maximum number of iterations
set.seed(523) # for reproducibility
param.mean <- sample(data_summary$mean_HR, m) # initial mean values for the latent classes, randomly sampled from the data
param.sd <- rep(sd(data_summary$mean_HR) / m, m) # initial standard deviation values for the latent classes, set to a fraction of the overall standard deviation of the data

for (itera in 0:maxitera) { # EM algorithm loop
  names(param.mean) <- paste("mean",1:m) # assign names to the mean parameters for better readability
  names(param.sd) <- paste("sd",1:m) # assign names to the standard deviation parameters for better readability
  print(param.mean) # print the current mean parameters for each latent class
  print(param.sd) # print the current standard deviation parameters for each latent class

  ## Expectation-Maximization - E-Step

  xx <- data_summary$mean_HR # data points (mean HR for each subject)
  densities <- sapply(1:m, function(i){dnorm(xx,mean=param.mean[i],sd=param.sd[i])}) # calculate the density of each data point under each latent class distribution
  relevance <- t(apply(densities, 1, function(l){l/sum(l)})) # calculate the relevance (responsibility) of each latent class for each data point by normalizing the densities

  ## Expectation-Maximization - M-Step crisp

  attrib <- t(apply(relevance, 1, function(l){replace(rep(0,m), which.max(l)[1], 1)})) # assign each data point to the latent class with the highest relevance (crisp assignment)
  param.mean.new <- sapply(1:m, function(j){mean(data_summary$mean_HR[attrib[,j]==1])}) # update the mean parameters for each latent class by calculating the average of the data points that are assigned to that class
  param.sd.new <- sapply(1:m, function(j){sd(data_summary$mean_HR[attrib[,j]==1])}) # update the standard deviation parameters for each latent class by calculating the standard deviation of the data points that are assigned to that class

  ## Expectation-Maximization - Convergence

  cat("Iteration",itera,"evolution:",dif <- sum(abs(param.mean.new - param.mean)), "\n") # calculate the difference between the new and old mean parameters to check for convergence
  if (dif < eps) break # if the difference is less than the convergence threshold, exit the loop
  param.mean <- param.mean.new # update the mean parameters for the next iteration
  param.sd <- param.sd.new # update the standard deviation parameters for the next iteration
}
```

### 3 latent classes

```{r}
# iteration 0
m <- 3 # number of latent classes
eps <- 0.01 # convergence error
itera <- 0 # iteration counter
maxitera <- 100 # maximum number of iterations
set.seed(523) # for reproducibility
param.mean <- sample(data_summary$mean_HR, m) # initial mean values for the latent classes, randomly sampled from the data
param.sd <- rep(sd(data_summary$mean_HR) / m, m) # initial standard deviation values for the latent classes, set to a fraction of the overall standard deviation of the data

for (itera in 0:maxitera) { # EM algorithm loop
  names(param.mean) <- paste("mean",1:m) # assign names to the mean parameters for better readability
  names(param.sd) <- paste("sd",1:m) # assign names to the standard deviation parameters for better readability
  print(param.mean) # print the current mean parameters for each latent class
  print(param.sd) # print the current standard deviation parameters for each latent class

  ## Expectation-Maximization - E-Step

  xx <- data_summary$mean_HR # data points (mean HR for each subject)
  densities <- sapply(1:m, function(i){dnorm(xx,mean=param.mean[i],sd=param.sd[i])}) # calculate the density of each data point under each latent class distribution
  relevance <- t(apply(densities, 1, function(l){l/sum(l)})) # calculate the relevance (responsibility) of each latent class for each data point by normalizing the densities

  ## Expectation-Maximization - M-Step crisp

  attrib <- t(apply(relevance, 1, function(l){replace(rep(0,m), which.max(l)[1], 1)})) # assign each data point to the latent class with the highest relevance (crisp assignment)
  param.mean.new <- sapply(1:m, function(j){mean(data_summary$mean_HR[attrib[,j]==1])}) # update the mean parameters for each latent class by calculating the average of the data points that are assigned to that class
  param.sd.new <- sapply(1:m, function(j){sd(data_summary$mean_HR[attrib[,j]==1])}) # update the standard deviation parameters for each latent class by calculating the standard deviation of the data points that are assigned to that class

  ## Expectation-Maximization - Convergence

  cat("Iteration",itera,"evolution:",dif <- sum(abs(param.mean.new - param.mean)), "\n") # calculate the difference between the new and old mean parameters to check for convergence
  if (dif < eps) break # if the difference is less than the convergence threshold, exit the loop
  param.mean <- param.mean.new # update the mean parameters for the next iteration
  param.sd <- param.sd.new # update the standard deviation parameters for the next iteration
}
```

## Using Mclust package to fit Gaussian mixture models with 2 and 3 latent classes. The `Mclust` function will automatically select the best model based on Bayesian Information Criterion, but we will specify the number of classes (G) to fit the models accordingly. We also set convergence criteria for the EM algorithm to ensure it runs efficiently.

```{r}
# Set convergence error and maximum number of iterations for EM algorithm
eps <- 0.01
itmax <- 100
# Apply EM algorithm with 2 latent classes
em_2 <- Mclust(data_summary$mean_HR, G = 2, control = emControl(eps=eps, itmax=itmax))
# Apply EM algorithm with 3 latent classes
em_3 <- Mclust(data_summary$mean_HR, G = 3, control = emControl(eps=eps, itmax=itmax))
```

# 4. Plot the densities for each of the latent classes.

# 5. Plot the mean HR as a function of each subject's age in which each point has a color corresponding to the class to which it belongs.

# 6. Generate 1000 data points at random according to a single distribution fitted to the original data.

# 7. Generate 1000 data points at random according to a mixture of distributions fitted using one of the previous EM computed in 3.
