---
title: "Assignment_3"
author: "João Silva, Marco Pereira, Mariana Ribeiro"
date: "2026-04-08"
output: pdf_document
---

# Work Description

## Simulation Overview

Simulate a population with **100,000 subjects**.

**Population Demographics:**

-   48% are male.

-   45% are 50 years old or older.

-   27% of the males smoke.

-   16% of the females smoke.

    *Note: [Risk factors are considered independent.]{.underline}*

**Risk Estimates:**

For this population, the following cumulative risks are estimated:

-   **5-year risk of disease:** 17.6%

    -   50.3% if a smoker \| 5.4% if a non-smoker

    -   22.1% if ≥50 years old \| 13.2% if \<50 years old

-   **10-year risk of any-cause mortality:** 11.5%

    -   22.9% if ≥50 years old \| 1.0% if \<50 years old

    -   12.3% if male \| 10.6% if female

## Data Structure Requirements

The resulting data frame must contain the following:

-   A column for each risk factor.

-   **Yearly risk of death:** i.e., the transition probability from alive to dead.

-   **Yearly risk of disease:** i.e., the transition probability from healthy to ill.

## Primary Objectives

Based on the simulation, answer the following:

1.  How many of the patients developed the disease after 10 years?

2.  How many of the patients died after 10 years?

## Load packages

```{r, echo=FALSE}

```

# Block 1: Initial Cohort Generation

The foundation of our model requires generating a baseline population of 100,000 individuals (Time 0). To ensure the strict reproducibility required by Monte Carlo simulations, we begin by defining a fixed seed. We generate the independent demographic variables (Gender and Age Group) using simple sampling with replacement, adhering to the marginal probabilities provided in the assignment (48% male; 45% aged 50 or older).

```{r}

# Reprodutibility and sample size
set.seed(9) 
N <- 100000

# Generation of Independent Variables
  # Gender - 48% male, 52% female
pop_sex <- sample(
  x = c("Male", "Female"), 
  size = N, 
  replace = TRUE, 
  prob = c(0.48, 0.52)
)
  # Age - 45% >=50, 55% <50.
pop_age <- sample(
  x = c(">=50", "<50"), 
  size = N, 
  replace = TRUE, 
  prob = c(0.45, 0.55)
)
```

Unlike age and gender, smoking status is a conditional variable. We cannot generate it blindly, as the probability of being a smoker is explicitly dependent on gender (27% for men versus 16% for women). To solve this, we allocate an empty vector and apply targeted sampling based on gender indices.

```{r}
# Generation of the Conditional Variable (Smoking)
   # 27% of the males smoke.
   # 16% of the females smoke.
# Initialise an empty vector
pop_smoke <- character(N)

# Identify indexes of men and women
idx_male <- which(pop_sex == "Male")
idx_female <- which(pop_sex == "Female")

# Simulate smoking ONLY for men (27% smokers)
pop_smoke[idx_male] <- sample(
  x = c("Smoker", "Non-Smoker"),
  size = length(idx_male),
  replace = TRUE,
  prob = c(0.27, 0.73)
)

# Simulate smoking ONLY for women (16% smokers)
pop_smoke[idx_female] <- sample(
  x = c("Smoker", "Non-Smoker"),
  size = length(idx_female),
  replace = TRUE,
  prob = c(0.16, 0.84)
)
```

We consolidate the generated vectors into a single data frame, immediately converting them to factors. We also pre-allocate empty columns for the annual risks, which will be populated in Block 2.

```{r}
# Building the data frame
cohort <- data.frame(
  ID = 1:N,
  Gender = factor(pop_sex, levels = c("Male", "Female")),
  AgeGroup = factor(pop_age, levels = c("<50", ">=50")),
  SmokingStatus = factor(pop_smoke, levels = c("Non-Smoker", "Smoker"))
)

# Added already required blank columns to be filled later
cohort$YearlyRisk_Disease <- NA
cohort$YearlyRisk_Death <- NA

# View dataframe
cohort
```

In large-scale stochastic simulations, we must empirically verify our generated data against our theoretical assumptions. The tables below confirm that our Monte Carlo generation successfully converges on the target proportions required by the assignment.

```{r}
# Validation (Quality Check)
print("--- Validation: Gender Distribution ---")
round(prop.table(table(cohort$Gender)) * 100, 2)

print("--- Validation: Age Distribution ---")
round(prop.table(table(cohort$AgeGroup)) * 100, 2)

print("--- Validation: Smoking by Gender ---")
round(prop.table(table(cohort$Gender, cohort$SmokingStatus), margin = 1) * 100, 2)

```

# Block 2: Risk Combination and Conversion to Annual Rates

Implementing the risk calculation logic is the most critical step in our model. We start with the base cumulative risks for disease and death, then apply multiplicative modifiers based on the individual's risk factors (smoking status, age

```{r}
# Risk Calculation Logic
# Base risks
base_risk_disease_5yr <- 0.176
base_risk_death_10yr <- 0.115
# Risk modifiers
risk_modifiers <- list(
  Disease = list(
    Smoking = c("Smoker" = 0.503, "Non-Smoker" = 0.054),
    AgeGroup = c(">=50" = 0.221, "<50" = 0.132)
  ),
  Death = list(
    AgeGroup = c(">=50" = 0.229, "<50" = 0.010),
    Sex = c("Male" = 0.123, "Female" = 0.106)
  )
)
```

The cumulative risks are calculated by multiplying the base risk with the appropriate modifiers for each individual's characteristics.

```{r}
# Calculate individual risks
cohort$Risk_Disease_5yr <- base_risk_disease_5yr * ifelse(
        cohort$SmokingStatus == "Smoker", 
        risk_modifiers$Disease$Smoking["Smoker"],
        risk_modifiers$Disease$Smoking["Non-Smoker"]
        ) * ifelse(
                cohort$AgeGroup == ">=50", 
                risk_modifiers$Disease$AgeGroup[">=50"],
                risk_modifiers$Disease$AgeGroup["<50"]
                )
cohort$Risk_Death_10yr <- base_risk_death_10yr *  ifelse(
        cohort$AgeGroup == ">=50", 
        risk_modifiers$Death$AgeGroup[">=50"],
        risk_modifiers$Death$AgeGroup["<50"]
        ) *  ifelse(
                cohort$Gender == "Male", 
                risk_modifiers$Death$Sex["Male"],
                risk_modifiers$Death$Sex["Female"]
                )
```

After obtaining the cumulative risks, we convert them to annual rates using the formula: $$\text{Yearly Risk} = -\frac{\log(1 - \text{Cumulative Risk})}{\text{Time Period}}$$ This conversion assumes a constant hazard rate over the specified time period, which is a common approach in Markov models to derive transition probabilities from cumulative risks

```{r}
# Convert cumulative risks to annual rates
cohort$YearlyRisk_Disease <- -log(1 - cohort$Risk_Disease_5yr) / 5
cohort$YearlyRisk_Death <- -log(1 - cohort$Risk_Death_10yr) / 10
```

# Block 3: Markov Chain and Results Analysis

# Conclusions
