---
title: "Assignment_3"
author: "João Silva, Marco Pereira, Mariana Ribeiro"
date: "2026-04-08"
output: pdf_document
---

# Work Description

## Simulation Overview

Simulate a population with **100,000 subjects**.

**Population Demographics:**

-   48% are male.

-   45% are 50 years old or older.

-   27% of the males smoke.

-   16% of the females smoke.

    *Note: [Risk factors are considered independent.]{.underline}*

**Risk Estimates:**

For this population, the following cumulative risks are estimated:

-   **5-year risk of disease:** 17.6%

    -   50.3% if a smoker \| 5.4% if a non-smoker

    -   22.1% if ≥50 years old \| 13.2% if \<50 years old

-   **10-year risk of any-cause mortality:** 11.5%

    -   22.9% if ≥50 years old \| 1.0% if \<50 years old

    -   12.3% if male \| 10.6% if female

## Data Structure Requirements

The resulting data frame must contain the following:

-   A column for each risk factor.

-   **Yearly risk of death:** i.e., the transition probability from alive to dead.

-   **Yearly risk of disease:** i.e., the transition probability from healthy to ill.

## Primary Objectives

Based on the simulation, answer the following:

1.  How many of the patients developed the disease after 10 years?

2.  How many of the patients died after 10 years?

## Load packages

```{r, echo=FALSE}

```

# Block 1: Initial Cohort Generation

The foundation of our model requires generating a baseline population of 100,000 individuals (Time 0). To ensure the strict reproducibility required by Monte Carlo simulations, we begin by defining a fixed seed.
We generate the independent demographic variables (Gender and Age Group) using simple sampling with replacement, adhering to the marginal probabilities provided in the assignment (48% male; 45% aged 50 or older).

```{r}

# Reprodutibility and sample size
set.seed(9) 
N <- 100000

# Generation of Independent Variables
  # Gender - 48% male, 52% female
pop_sex <- sample(
  x = c("Male", "Female"), 
  size = N, 
  replace = TRUE, 
  prob = c(0.48, 0.52)
)
  # Age - 45% >=50, 55% <50.
pop_age <- sample(
  x = c(">=50", "<50"), 
  size = N, 
  replace = TRUE, 
  prob = c(0.45, 0.55)
)
```

Unlike age and gender, smoking status is a conditional variable. We cannot generate it blindly, as the probability of being a smoker is explicitly dependent on gender (27% for men versus 16% for women). To solve this, we allocate an empty vector and apply targeted sampling based on gender indices.

```{r}
# Generation of the Conditional Variable (Smoking)
   # 27% of the males smoke.
   # 16% of the females smoke.
# Initialise an empty vector
pop_smoke <- character(N)

# Identify indexes of men and women
idx_male <- which(pop_sex == "Male")
idx_female <- which(pop_sex == "Female")

# Simulate smoking ONLY for men (27% smokers)
pop_smoke[idx_male] <- sample(
  x = c("Smoker", "Non-Smoker"),
  size = length(idx_male),
  replace = TRUE,
  prob = c(0.27, 0.73)
)

# Simulate smoking ONLY for women (16% smokers)
pop_smoke[idx_female] <- sample(
  x = c("Smoker", "Non-Smoker"),
  size = length(idx_female),
  replace = TRUE,
  prob = c(0.16, 0.84)
)
```

We consolidate the generated vectors into a single data frame, immediately converting them to factors. We also pre-allocate empty columns for the annual risks, which will be populated in Block 2.

```{r}
# Building the data frame
cohort <- data.frame(
  ID = 1:N,
  Gender = factor(pop_sex, levels = c("Male", "Female")),
  AgeGroup = factor(pop_age, levels = c("<50", ">=50")),
  SmokingStatus = factor(pop_smoke, levels = c("Non-Smoker", "Smoker"))
)

# Added already required blank columns to be filled later
cohort$YearlyRisk_Disease <- NA
cohort$YearlyRisk_Death <- NA

# View dataframe
cohort
```

In large-scale stochastic simulations, we must empirically verify our generated data against our theoretical assumptions. The tables below confirm that our Monte Carlo generation successfully converges on the target proportions required by the assignment.

```{r}
# Validation (Quality Check)
print("--- Validation: Gender Distribution ---")
round(prop.table(table(cohort$Gender)) * 100, 2)

print("--- Validation: Age Distribution ---")
round(prop.table(table(cohort$AgeGroup)) * 100, 2)

print("--- Validation: Smoking by Gender ---")
round(prop.table(table(cohort$Gender, cohort$SmokingStatus), margin = 1) * 100, 2)

```

# Block 2: Risk Combination and Conversion to Annual Rates

Since the provided risks are cumulative over 5 or 10 years, we need to convert them into annual transition probabilities. We will use the formula for converting cumulative probability to a rate, and then back to an annual probability. This process is done separately for disease risk and death risk, based on the relevant risk factors. The assumption of independence allows us to simply sum the rates for each risk factor to get a total rate for each individual, which we then convert back to an annual probability.

## Disease Risk

```{r}
# Function to convert probability to rate
prob_to_rate <- function(p, t) { -log(1 - p) / t }

# Convert 5-year probabilities to rates for each risk factor
r_smoker <- prob_to_rate(0.503, 5)
r_non_smoker <- prob_to_rate(0.054, 5)
r_old <- prob_to_rate(0.221, 5)
r_young <- prob_to_rate(0.132, 5)

# Convert rates back to yearly probabilities
cohort$YearlyRisk_Disease_Smoker <- 1 - exp(-r_smoker * 1)
cohort$YearlyRisk_Disease_NonSmoker <- 1 - exp(-r_non_smoker * 1)
cohort$YearlyRisk_Disease_Old <- 1 - exp(-r_old * 1)
cohort$YearlyRisk_Disease_Young <- 1 - exp(-r_young * 1)
```

## Death Risk

```{r}
# Convert 10-year probabilities to rates for each risk factor
r_old_death <- prob_to_rate(0.229, 10)
r_young_death <- prob_to_rate(0.010, 10)
r_male_death <- prob_to_rate(0.123, 10)
r_female_death <- prob_to_rate(0.106, 10)

# Convert total rates back to yearly probabilities
cohort$YearlyRisk_Death_Old <- 1 - exp(-r_old_death * 1)
cohort$YearlyRisk_Death_Young <- 1 - exp(-r_young_death * 1)
cohort$YearlyRisk_Death_Male <- 1 - exp(-r_male_death * 1)
cohort$YearlyRisk_Death_Female <- 1 - exp(-r_female_death * 1)
```

## Total Risk Calculation

To check the answers obtained in the simulation we can calculate the total risk of disease in 10 years for the population and use it and the 10-year risk of any-cause mortality as a benchmark.

```{r}
# Convert 5-year probability of risk of disse to rate
r_disease <- prob_to_rate(0.176, 5)

# Convert year rates back to 10-year probability
total_risk_disease <- 1 - exp(-r_disease * 10)

# Calculate how many of the patients developed the disease after 10 years.
num_disease <- round(total_risk_disease * N)
num_disease

# Calculate how many of the patients died after 10 years.
total_risk_death <- 0.115
num_death <- round(total_risk_death * N)
num_death
```

Based on the cumulative probabilities, the answers are the following:

1.  How many of the patients developed the disease after 10 years?

After 10 years the number of patients that developed the disease is `r num_disease`.

2.  How many of the patients died after 10 years?

After 10 years the number of patients that died is `r num_death`.

# Block 3: Markov Chain and Results Analysis

```{r}

# 1. MORTALITY
cohort$r_death <- (ifelse(cohort$AgeGroup == ">= 50", 
                        r_old_death, 
                        r_young_death)
                        +
                  ifelse(cohort$Gender == "Male",
                        r_male_death,
                        r_female_death))

# Convert total death rate back to annual probability
cohort$p_death_annual <- 1 - exp(-cohort$r_death)


# 2. DISEASE
cohort$r_disease <- (ifelse(cohort$SmokingStatus == "Smoker", 
                            r_smoker, 
                            r_non_smoker)
                            +
                     ifelse(cohort$Gender == "Male",
                        r_male_death,
                        r_female_death))

# Convert total disease rate back to annual probability
cohort$p_disease_annual <- 1 - exp(-cohort$r_disease)
```

```{r}
# Markov Simulation

# 1. Initialization of the population state
# All 100,000 subjects start in the "Healthy" state
n_subjects <- nrow(cohort)
status <- rep("Healthy", n_subjects)

# 2. Iterative Markov Loop (Year 1 to 10)
for (yr in 1:10) {
  
  # Step A: Identify subjects eligible for mortality (anyone not already dead)
  # In a Markov model, the transition depends on the current state
  alive_idx <- which(status != "Dead")
  
  # Step B: Simulate Mortality (Transition to 'Dead' state)
  # Generate a random number [0,1] for each alive subject
  death_random_draw <- runif(length(alive_idx))
  
  # If the random draw is less than the specific annual risk, the subject dies
  died_this_year <- death_random_draw < cohort$p_death_annual[alive_idx]
  
  # Update the status of those who transitioned to the absorbing state "Dead"
  status[alive_idx[died_this_year]] <- "Dead"
  
  
  # Step C: Simulate Disease (Transition to 'Ill' state)
  # Only those who are "Healthy" AND survived the mortality check can transition
  # This handles the competing risk: death "wins" over disease if both occur
  healthy_survivors_idx <- which(status == "Healthy")
  
  disease_random_draw <- runif(length(healthy_survivors_idx))
  
  # If the draw is less than the risk, the subject moves from Healthy -> Ill
  developed_disease_this_year <- disease_random_draw < cohort$p_disease_annual[healthy_survivors_idx]
  
  # Update the status to "Ill"
  status[healthy_survivors_idx[developed_disease_this_year]] <- "Ill"
}

# 3. Final Outcome Extraction
# Count final states after 10 years
final_deaths <- sum(status == "Dead")
final_disease_cases <- sum(status == "Ill")

# Output the results for validation
cat("Simulation Results after 10 years:\n")
cat("- Total Patients who Died:", final_deaths, "\n")
cat("- Total Patients who Developed Disease (and stayed alive):", final_disease_cases, "\n")
```


# Conclusions


```{r}
# The discrepancy between the simulated results and the expected values based on cumulative probabilities can be attributed to several factors:
# 1. Stochastic Variability: Monte Carlo simulations inherently involve randomness. Even with a large sample size of 100,000, there can be variability in the outcomes due to chance. Running multiple iterations of the simulation and averaging results can help mitigate this.
# 2. Independence Assumption: The assumption that risk factors are independent may not hold true in real-world scenarios. If there are interactions between risk factors (e.g., smoking and age), the combined effect on disease and mortality risk may differ from the simple additive model used in the simulation.
# 3. Transition Probability Estimation: The conversion of cumulative probabilities to annual rates and back to probabilities may introduce approximation errors, especially if the underlying risk does not follow a simple exponential distribution. The actual risk over time may be more complex than the model assumes. 

```